---
title: "Dimensional autism"
subtitle: "- Taxometric analysis"
author: "enrico toffalini @psicostat @m&llab"
date: "January 16th, 2024"
output: html_document
---
\raggedright
\break\break

```{r setup-knitr, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 

```

The present set of simulation studies suggests that taxometric analysis works well in detecting the dimensional structure of a population when the sample is representative the population (Study 1). In incurs false positive result (incorrectly favoring a category structure) when diagnosed cases are deliberately oversampled (Study 3). Somehow surprisingly, it also incur false positive results when the sample is representative of a portion of the population (upper tail; Study 2).

--------------------------------------------
--------------------------------------------

### Initial general setup

```{r setup-general} 
rm(list=ls())
library(RTaxometrics)
library(effsize)

set.seed(0)
```

--------------------------------------------
--------------------------------------------

## GENERATE A DIMENSIONAL POPULATION

Let's create a large population of 10^6^ individuals. In it, there is an "autistic trait" varying dimensionally and continuously, distributed as a standard Normal. 

```{r, cache=T}
Npop = 1e6
realAutism = rnorm(Npop,0,1)

```

Since this trait is *latent*, it cannot be observed. So, let's assume that all individuals undergo an assessment measuring 3 diagnostic indexes that reflect the real trait with some noise. They are simulated as follows. The correlation between each pair of indices turns out to be *r* = 0.80.

```{r, cache=T}
index1 = scale(realAutism + rnorm(Npop,0,0.5))
index2 = scale(realAutism + rnorm(Npop,0,0.5))
index3 = scale(realAutism + rnorm(Npop,0,0.5))
round(cor(cbind(index1,index2,index3)),2)
```

Now let's put the indexes in a data frame, and let's simulate the diagnostic process. Diagnosis is done dimensionally: if the combined mean score in the 3 indexes is in the top 5% quantile, the individual gets a diagnosis ("2", while nondiagnosed individuals are "1").

```{r, cache=T}
pop = data.frame(index1,index2,index3)
pop$indexTot = rowMeans(pop[,c("index1","index2","index3")])
pop$diagn = ifelse(pop$indexTot>quantile(pop$indexTot,.95),2,1)
```

Finally, let's simulate a set of 5 indicators (x1 to x5) that are affected by the latent autistic trait, but more loosely than the previous 3 indexes (i.e., there is more noise here). The 5 indicators correlate *r* = 0.31 with each other.


```{r, cache=T}
pop$x1 = scale(realAutism + rnorm(Npop,0,1.5))
pop$x2 = scale(realAutism + rnorm(Npop,0,1.5))
pop$x3 = scale(realAutism + rnorm(Npop,0,1.5))
pop$x4 = scale(realAutism + rnorm(Npop,0,1.5))
pop$x5 = scale(realAutism + rnorm(Npop,0,1.5))
round(cor(pop[,c("x1","x2","x3","x4","x5")]),2)
```

Finally, let's have a look at how largely these indicators differ between individuals diagnosed vs nondiagnosed with autism. Let's use x1 (the other are identically related to the autistic trait). 

```{r, cache=T}
(cd = cohen.d(pop$x1 ~ pop$diagn))
```

The Cohen's d turn out to be *d* = `r round(cd$estimate,2)`.

--------------------------------------------
--------------------------------------------

## STUDY 1: DRAW A REPRESENTATIVE SAMPLE

In Study 1 we draw a large representative sample (N = 20000) from the general population and we run taxometric analysis on it, using the five indicators x1 to x5.

First, we randomly draw the sample from the population.

```{r study-1-draw-sample, cache=T}
N = 20000

selected = sample(1:nrow(pop),N)
df = pop[selected,]
```

Second, we look at how many "cases" and how many "controls" were drawn. They are pretty much in line with the overall population (where diagnosed children are exactly 5%).

```{r study-1-see-cases, cache=T}
table(df$diagn)
```

Finally, let's run the taxometric analysis.

```{r study-1-run-taxometrics, cache=T}

(rt_study1 = RunTaxometrics(df[,c("x1","x2","x3","x4","x5","diagn")]))
```

**So, in Study 1 the resulting mean CCFI is `r round(rt_study1$CCFI.mean,2)`, thus favoring a dimensional structure (which is correct, because a dimensional structure is the ground truth).**

--------------------------------------------
--------------------------------------------

## STUDY 2: DRAW A REPRESENTATIVE SAMPLE OF THE UPPER TAIL OF THE CURVE (z > +0.5)

In Study 2 we draw a large sample (N = 20000) that is representative of the upper half of the population only (i.e., where the combined index1, index2, index3, is z > +0.5) and once again we run taxometric analysis on it, using the five indicators x1 to x5.

First, let's randomly sample the cases. Given the inclusion criterion, the share of diagnosed cases in the sample is now obviously larger than in Study 1.

```{r study-2-draw-sample, cache=T}
N = 20000

pop$zscore = scale(rowSums(pop[,c("index1","index2","index3")]))
pop_upper = pop[pop$zscore > +0.5,]

selected = sample(1:nrow(pop_upper),N)
df = pop_upper[selected,]

table(df$diagn)
```

Now let's run the taxometric analysis.

```{r study-2-run-taxometrics, cache=T}
(rt_study2 = RunTaxometrics(df[,c("x1","x2","x3","x4","x5","diagn")]))
```

**So, in Study 2 the resulting mean CCFI is `r round(rt_study2$CCFI.mean,2)`, thus incorrectly favoring a categorical structure.**

--------------------------------------------
--------------------------------------------

## STUDY 3: DRAW A SAMPLE WHERE PREVIOUSLY IDENTIFIED DIAGNOSED CASES ARE DELIBERATELY OVERSAMPLED (AND THEY REPRESENT ABOUT 50% OF THE OVERALL TESTED SAMPLE)

In Study 3 we draw a large sample (N = 20000) that is deliberately unbalanced: individuals with a diagnosis (`diagn=="2"`) are sample separately from those without a diagnosis (`diagn==1`). To do so, a much heavier probability sampling weight is used for diagnosed individuals. The weight is chosen to ensure that about 50% of the overall sample will be represented by individuals with a diagnosis.

First, let's randomly sample the cases. Given the inclusion criteria, the share of diagnosed cases in the sample is about half.

```{r study-3-draw-sample, cache=T}
N = 20000
prob = ifelse(pop$diagn==2,20,1)

selected = sample(1:nrow(pop),N,prob=prob)

df = pop[selected,]
table(df$diagn)
```


Now let's run the taxometric analysis.

```{r study-3-run-taxometrics, cache=T}
(rt_study3 = RunTaxometrics(df[,c("x1","x2","x3","x4","x5","diagn")]))
```

**So, in Study 3 the resulting mean CCFI is `r round(rt_study3$CCFI.mean,2)`, thus incorrectly favoring a categorical structure with very strong evidence.**


--------------------------------------------
--------------------------------------------

## CONCLUSIONS

As said above, results suggest that taxometric analysis works well in detecting the dimensional structure of a population when the sample is representative the population (Study 1). In incurs false positive result (incorrectly favoring a category structure) when diagnosed cases are deliberately oversampled (Study 3). Somehow surprisingly, it also incur false positive results when the sample is representative of a portion of the population (upper tail; Study 2). But of course, I might have got something wrong!

--------------------------------------------
--------------------------------------------

